Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-02-03 manu_draft.Rmd
discussion revision from JH done

@@

The approach developed here can be used to estimate seasonal averages with uncertainty in water quality parameters and evaluate trends in these seasonal changes across years using full propagation of error estimates across methods.  This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an "optimal" level of smoothing with the mgcv R package that uses generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17]. Importantly, specifying an upper limit for the number of knots that can potentially be used to fit different smoothers is critical to this approach.  A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risk inflated Type I error rates. We further demonstrated that models with different structures in the smoothed components, such as separate smoothers for continuous year and day of year, can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package (Figure \@ref(fig:modsumfig)).  This comparison demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends over time. Although models with different structures allocate variance components of the response variable differently, this is a trivial issue for the second and third stage of analyses to estimate significance of trends developed herein.

@@

Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  These comparisons are analogous to the example in the previous paragraph where a common approach prior to trend analysis is to aggregate raw observations.  Our assessment of trends using linear regression applied to seasonal averages from the raw observations is effectively similar to averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.  This is particularly problematic if sampling effort varies substantially between years.  Averages may be skewed in a particular direction if annual estimates are based on a handful of observations from select months (e.g., summer only)[@Fouquet12].  The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages. More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses [@Sera19].  As a result, trend assessments from mixed-meta regression should be considered an accurate representation of the actual trend within the limits of the raw observations.

The underlying cross-validation methods used by GAMs in the mgcv package also reduce the decisions required by an analyst that may be necessary for the implementation of alternative trend assessment methods.  For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing [@Hirsch10;@Wan17].  Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, leading to potential under- or over-smoothing of fitted results to the raw observations.  Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced by process or observation error present in the raw data.  However, we acknowledge that the third stage analyses that require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.  Although there are undoubtedly many scenarios where years of interest can be chosen objectively (e.g., regulatory compliance periods, time since management intervention, etc.), a more general question of "when" changes occur independent of user decisions is also important to address.  For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?  There may be additional methods to develop using objective criteria to more accurately identify inflection points or other important periods where changes occur independent of a user choice.

The trend assessment approach also does not address causality in the observed trends.  In this regard, the approach is similar to other trend assessment methods where the focus is understanding direction, magnitude, and confidence in changes in water quality variables over time. However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.  This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.  An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends [@Wood02;@Zuur09]. In such cases, considerations of model structure can have direct implications on conclusions regarding potential causality.  As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.  Our goal herein was to describe chlorophyll changes relative to time where the predictors were variations on a general theme (e.g., season vs. year), whereas using GAMs with different predictors assess potential causality is a different application with alternative goals.  Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to assess different questions.

Finally, the evaluation of trends for alternative water quality variables in addition to chlorophyll is a trivial but logical extension of the methods proposed in this study.  The long-term monitoring program maintained by USGS includes multiple parameters in addition to chlorophyll that can provide additional context into broader water quality trends in South Bay [@Cloern16;@Schraga20]. These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.  Chlorophyll measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].  We provide an open-source R package to facilitate these analyses (see supplement) that includes all methods described in this manuscript.

@@

The wqtrends R package includes all methods described in this manuscript.  The package is available for download at https://tbep-tech.github.io/wqtrends.  A full vignette describing installation and use is also available in the link provided.  Please cite this manuscript in applications using the R package.

@@

We thank the staff of the US Geological Survey that collect and maintain long-term monitoring data in San Francisco Bay.  This work benefited from discussions with the San Francisco Bay Nutrient Technical Workgroup and Steering Committee.

@@

  # plots
  p1 <- ggplot(data = tomod1, aes(x = yr, y = bt_avg)) +
    geom_point(colour = 'deepskyblue3') +
    geom_ribbon(data = toplo1, aes(ymin = bt_lwr, ymax = bt_upr), fill = 'grey', alpha = 0.4) +
    geom_line(data = toplo1, color = 'grey') +
    theme_minimal(base_family = 'serif') +
    coord_cartesian(ylim = ylim, xlim = c(1990, 2020)) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank()
    ) +
    labs(
      title = ttl,
      subtitle = paste('Obesrved regression on raw data,', pval1),
      y = NULL
    )

  p2 <- ggplot(data = avgseas, aes(x = yr, y = bt_avg)) +
    geom_point(colour = 'deepskyblue3') +
    geom_ribbon(data = toplo2, aes(ymin = bt_lwr, ymax = bt_upr), fill = 'lightgreen', alpha = 0.4) +
    geom_line(data = toplo2, color = 'green') +
    theme_minimal(base_family = 'serif') +
    coord_cartesian(ylim = ylim, xlim = c(1990, 2020)) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank()
    ) +
    labs(
      subtitle = paste('Average regression on GAM estimates,', pval2),
      y = ylab
    )

@@

p <- (p1 |  p2) + plot_layout(ncol = 2)
png('figs/trndcmpex.png', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at two example stations. The "observed" trend model is based on a linear fit to the annual averages of chlorophyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM model.  Regressions in each plot are fit through the seasonal estimates indicated in the plot titles.'}
knitr::include_graphics('figs/trndcmpex.png')
```

>>>>

### wqtrends R package

The wqtrends R package developed by the authors includes all methods described above.  The package is available for download at https://tbep-tech.github.io/wqtrends.  A full vignette describing installation and use is also available in the link provided.

@@

@@

Our results demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends over time. This is an important conclusion that can provide guidance on how GAMs should be used to model time series from long-term water quality monitoring data.  Models with separate smoothers for continuous year and day of year can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package (Figure \@ref(fig:modsumfig)).  This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an optimal level of smoothing with the mgcv R package using generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17]. Specifying an upper limit for the number of knots that can potentially be used to fit different smoothers is critical to this approach.  A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risks incorrect conclusions from using these models with the second and third stage of analyses.

Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  These comparisons demonstrate a common approach where raw observations may be aggregated prior to use of more conventional trend analysis methods.  Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.  A false negative result (Type II) was shown in Figure \@ref(fig:trndcmpex)a where a trend was not shown for the OLS regression results, but was observed for the mixed-meta regression results.  Conversely, a false positive result (type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, whereas the mixed-meta results did not.  The potential danger of inflated error rates was further demonstrated at a larger spatial scale across all stations in Figure \@ref(fig:trndcmp).

Variability of sampling effort among years is a common characteristic of environmental monitoring data, which further increases the potential for incorrect conclusions when existing methods have been used. Averages may be skewed in a particular direction if annual estimates are based on a handful of observations from select months [e.g., summer only, @Fouquet12].  The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages. More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses [@Sera19].  As a result, trend assessments from mixed-meta regression are accurate for data with unequal sampling effort where other methods may fail by providing estimates that are biased or have inaccurate or incomplete estimates of uncertainty.

The underlying cross-validation methods used by GAMs in the mgcv package also reduce the decisions required by an analyst that may be necessary for the implementation of alternative trend assessment methods.  For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing [@Hirsch10;@Wan17].  Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, producing fitted results that are arbitraily under- or over-smoothed to the raw observations.  This is especially problematic when for policy analysis or regulatory decisions if the results change based on arbitrary decisions of the analyst.  Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced by process or observation error present in the raw data.

## Future work

Additional work could be conducted to further strengthen the validity of conclusions based on trends from mixed-meta regression analyses applied to the seasonal GAM averages.  We acknowledge that the third stage analyses require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.  Although there are undoubtedly many scenarios where years of interest can be chosen objectively (e.g., regulatory compliance periods, time since management intervention, etc.), a more general question of "when" changes occur independent of user decisions is also important to address.  For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?  There may be additional methods to develop using objective criteria to more accurately identify inflection points or other important periods where changes occur independent of a user choice.

The trend assessment approach does not address causality in the observed trends.  In this regard, the approach is like other trend assessment methods where the focus is understanding direction, magnitude, and confidence in changes in water quality variables over time. However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.  This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.  An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends [@Wood02;@Zuur09]. In such cases, considerations of model structure can have direct implications on conclusions regarding potential causality.  As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.  Our goal herein was to describe chlorophyll changes relative to time where the predictors were variations on a general theme (e.g., season vs. year), whereas using GAMs with different predictors assess potential causality is a different application with alternative goals.  Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to assess different questions.

Finally, the evaluation of trends for alternative water quality variables in addition to chlorophyll is a simple and logical extension of the methods proposed in this study.  The long-term monitoring program maintained by USGS includes multiple parameters in addition to chlorophyll that can provide additional context into broader water quality trends in South Bay [@Cloern16;@Schraga20]. These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.  Chlorophyll measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].  The open-source R package developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.

@@

@@

We thank the staff of the US Geological Survey that collect and maintain long-term monitoring data in San Francisco Bay.  This work benefited from discussions with the San Francisco Bay Nutrient Technical Workgroup and Steering Committee.  We thank James D. Hagy III for reviewing an earlier draft of this manuscript.

@@

  # plots
  p1 <- ggplot(data = tomod1, aes(x = yr, y = bt_avg)) +
    geom_point(colour = 'deepskyblue3') +
    geom_ribbon(data = toplo1, aes(ymin = bt_lwr, ymax = bt_upr), fill = 'grey', alpha = 0.4) +
    geom_line(data = toplo1, color = 'grey') +
    theme_minimal(base_family = 'serif') +
    coord_cartesian(ylim = ylim, xlim = c(1990, 2020)) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank()
    ) +
    labs(
      title = ttl,
      subtitle = paste('OLS regression on raw data,', pval1),
      y = NULL
    )

  p2 <- ggplot(data = avgseas, aes(x = yr, y = bt_avg)) +
    geom_point(colour = 'deepskyblue3') +
    geom_ribbon(data = toplo2, aes(ymin = bt_lwr, ymax = bt_upr), fill = 'lightgreen', alpha = 0.4) +
    geom_line(data = toplo2, color = 'green') +
    theme_minimal(base_family = 'serif') +
    coord_cartesian(ylim = ylim, xlim = c(1990, 2020)) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank()
    ) +
    labs(
      subtitle = paste('OLS regression on GAM estimates,', pval2),
      y = ylab
    )

@@

p <- (p1 |  p2) + plot_layout(ncol = 2)
png('figs/trndcmpex.png', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at two example stations. The first row shows OLS (ordinary least squares) regression applied to annual averages of chlorophyll from the observed data, the second row shows OLS regression applied to annual averages of chlorophyll from the GAM, and the third row shows mixed-meta regression applied to the annual averages of chlorophyll from the GAM.  Regressions in each plot are fit through the seasonal estimates indicated in the plot titles for a specified year range.'}
knitr::include_graphics('figs/trndcmpex.png')
```
